---
title: "Imprisonment Without Parole: Understanding America's Complex History of 
Incarceration"
author: "Richard Tran"
date: '2022-11-18'
output: html_document
---

```{r echo = FALSE, eval = TRUE, message = FALSE, include = FALSE}
source("../source/a4-helpers.R")
source("../source/analysis.R")
library(english)
library(knitr)
library(kableExtra)
```

## Abstract 
[TODO: Include a short abstract (about 50 words)]

## Introduction 
From 1970 to 2014, the United States experienced an almost four-fold increase in
its national prison population. The reasons for imprisonment abound; people
often face pretrial detention, where they are sent to jail to await trial,
and prison overcrowding uses local jails to accept transfers
from state and federal prisons. The Vera Institute's research shows that
although the national jail admission rate has decreased in recent years, the
active growth in the daily jail population suggests that the average length of
stay has increased from its original rate of about nine days in 1970. However,
the Vera Institute also reveals an interesting incarceration statistic: local
jail populations in small and mid-sized counties have outpaced the growth rate
of those in larger counties, contrary to the preconception that the opposite
might be true. This means that the issue of mass incarceration is not
endemic to a particular region, but rather, it's a national issue. In addition,
the wide variation in incarceration rates across similarly situated counties
suggests that incarceration rates are rooted in historic social disparities
(Subramanian, Henrichson, and Kang-Brown 7-12).

Central to this investigation are the **direct stakeholders** of this data:
that is, the millions of people who are annually sentenced to legal confinement
by the American justice system. Addressing the causes for national
jail and prison growth requires open acknowledgement of the people who are most
disproportionately incarcerated, so naturally, _freedom_ and _liberation_ are
values that are closely connected to this data. Borrowing from the phrase,
"three strikes and you're out," which roughly describes how criminals convicted
of their third felony face life in prison, individuals classified as repeat
offenders are sentenced to extended jail time. However, this only works to
increase the daily prison population. Likewise, innocent people under pretrial
detention are highly vulnerable to developing mental health problems while in
confinement, in large part due to the hostile environment of jails and prisons.
Both of these examples are only several of many which illustrate how the current
judicial system works against those wrongly accused. It is by no means a
coincidence, then, that external groups such as the American Legislative
Exchange Council (ALEC) or Corrections Corporation of America (CCA, now
CoreCivic), which stand to gain from the continued operation of jails and
prisons, are some of their strongest backers (Duvernay and Moran). For this
reason, _democracy_ and _accountability_ are inherent to an analysis of this
data. Based on our findings, we must advocate for policy changes that hold
policymakers accountable for not only _who gets incarcerated_—be
it convicted criminals or troubled immigrants and other individuals awaiting
trial—but also _why_ and _for how long_ they remain detained.

Lastly, the data produced by the Vera Institute is not just a collection
of statistics: it is a reflection of how failures in public policy have impacted
human lives, a story of how confinement has shaped the lives, families,
and communities of those subjected to time behind bars. Every person facing
imprisonment brings with them a unique set of circumstances, and so extra care
must be taken to understand how their background and identity could have played
a factor in their incarceration. In other words, We must take initiative
to analyze the data to shed light on not only the root causes of American mass
incarceration, but also determine what steps the **indirect stakeholders** of
this data, which includes whole communities, local support groups,
organizations, systems designers, and policymakers, should take to resolve
disparities in prison demographics. Rather than pouring economic resources into
the construction of expensive prisons—which only works to perpetuate gender and
racial disparities in incarceration—policymakers should consider reinvesting
in communities and pursuing criminal justice reform. To this end, the analysis
in this report will highlight existing problem areas that could serve as a
starting point for further discussion.

In exploring the Vera Institute's data, the following questions will be
addressed:

* How does the number of people in pretrial detention in 2014 compare to
the 1970?
* When was the highest number of people in pretrial detention recorded?
* Which counties had the highest pretrial detention rates in 2014, and what were
their relative sizes?
* What was the average jail population rate of African Americans across the
United States in 2014?
* What was the average jail population rate of White Americans across the
United States in 2014?
* Which counties had the highest average prison population from 2000 to
2014?

## Data Summary 

```{r format_vars, echo = FALSE}
num_pretrial_diff <- prettyNum(num_pretrial_diff, big.mark = ",")

top_pretrial_counties_str <- paste0("* ",
                                   pull(top_pretrial_counties_2014, location),
                                   collapse = "\n")

num_rural <- top_pretrial_counties_2014 %>%
  filter(urbanicity == "rural") %>%
  nrow()

if (num_rural != 1) {
  num_rural <- paste(as.character(english(num_rural)), "were")
} else {
  num_rural <- paste(as.character(english(num_rural)), "was")
}

avg_aa_jail_rate <- round(avg_aa_jail_rate)

avg_white_jail_rate <- round(avg_white_jail_rate)
```

According to the data provided by the Vera Institute, the difference in the
national pretrial jail population count in 1970 and the same calculation in
2014 was `r num_pretrial_diff`. This represents an almost five-fold increase in
the number of people in jail awaiting trial since 1970. Even after accounting
for population increases, this information alone is staggering. The contrast
between the 1970 and 2014 pretrial jail population counts emphasizes some of
the outstanding issues with the current justice system, including individuals'
right to a "speedy trial."

The year `r year_highest_pretrial_pop` saw the highest number of people in
pretrial detainment. Connecting to the previous analysis on pretrial jail
population counts, this implies that the number of people facing
pretrial detention is only compounding with each year. Evidently, some form of
action must be taken on behalf of policymakers to remedy this issue. Pretrial
detention is largely counterproductive to correcting criminal behavior,
especially in cases where individuals are wrongfully accused.

The ten counties with the highest pretrial detention rates in 2014 were as
follows:

`r top_pretrial_counties_str`.

Of the above counties, `r num_rural` designated as rural counties. This means
that the counties with the highest pretrial detention rates were of rural/small
size, which disproves the claim that the highest pretrial detention rates would
be found in the counties with the greatest populations. The fact that smaller
counties have observed higher pretrial detention rates demonstrates that the
problem of mass incarceration is not restricted to the largest hubs of America,
but rather, the issue is of national scale.

The data uncovers some of the racial disparities related to incarceration in
the United States. With reference to 2014, the average jail population rate of
black Americans was about `r avg_aa_jail_rate` per one hundred thousand
people, and yet the average jail population rate for white Americans in the
same year was only about `r avg_white_jail_rate`. Viewed differently,
white Americans were incarcerated at a rate fives times less than that of black
Americans in 2014. That black Americans disproportionately represent
the US jail population begs the question of whether or not racial biases
in policing have played a role in the increased arrests of members of
this racial group.

From 2000 to 2014, the following counties observed the highest average prison
population counts and rates:

```{r county_prison_pop_tables, echo = FALSE}
kable(counties_highest_avg_prison_pop,
      caption = "Top Counties by Prison Population") %>%
  kable_styling(full_width = FALSE,
                position = "float_left")
kable(counties_highest_avg_prison_rate,
      caption = "Top Counties by Prison Population Rate per 100k People") %>%
  kable_styling(full_width = FALSE,
                position = "right")
```

While the top five counties by average prison population size are located in
urban areas, the top five counties by average prison population rates are marked
by rural, small/mid status. This observation bears resemblance to the earlier
examination of the counties with the highest pretrial detention rates in 2014.
In both cases, counties of rural to small status were found to have higher
incarceration rates than counties of larger population, reinforcing the fact
that incarceration rates are not simply a product of county population size.
The issue of mass incarceration deserves recognition by agencies of all levels,
including local communities whose lack of legal resources may cause the people
living there to be underrepresented.


[TODO ... other sections ... See Canvas for details]

_All the very best with your work!_

plot of all counties and the group with highest incarceration for each county

approach
communciating ifnroamtion through the absence of data
lack thereof


## References
* Ava Duvernay, and Jason Moran. _13TH_. USA, 2016.
* Ram Subramanian, Christian Henrichson, and Jacob Kang-Brown.
_In Our Own Backyard: Confronting Growth and Disparities in
American Jails_. New York, NY: Vera Institute of Justice, 2015.